Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-08-06 docs/papers/shallow_performance/shallow_perf.tex
shallow paper: Correct the P8 clock speed in the text
fleshed out the psykal section and the conclusions

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)}
%
% author names and IEEE memberships
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{R.~W.~Ford, M.~Modani, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
\thanks{M.~Modani is with IBM India.}
}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
The abstract goes here.
\end{abstract}

@@

\subsection{The \psykal Approach}
Proliferation of lightweight cores is resulting in increased
complexity in the programming models required to achieve performance.
Combine with the complexity of e.g. a model simulating circulation in
the global ocean. Significant challenge in software engineering.
Propose an approach for meeting this challenge by separating the
computational science demands from the {\it e.g.} oceanographic
demands.

\begin{figure}
\centering
\caption{The \psykal Approach.}
\end{figure}

While encapsulation makes it easier to produce more robust software,
it can often come with a performance penalty. In this work we
investigate in detail the hardware and compiler dependencies of this
penalty and the steps required to remove it.

\subsection{The `Shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  In describing
our work, we term the version of the Shallow program obtained at the
beginning of this project the `original' version.

@@

{\bf Auto-generated BC code by reasoning about meta-data. But then,
  that brings us on to meta-data...}

@@

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
'chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
2~GHz.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
IBM Power 8      &       &             &         & 15.1.2     \\
\hline
\end{tabular}
\end{table}

@@

@@

The take-home message of all this is that while we can employ a
separation of concerns and recover performance, doing so is not
straightforward. Determining how best to optimise even a simple code
such as Shallow is highly compiler- and CPU-dependent. Decisions that
are good on one platform may be bad for another and if these are
written into the source code then they will accumulate over time and
will almost certainly result in a code that does not perform optimally
on any system.

@@

We have investigated the application of the \psykal separation of
concerns approach to the domain of finite-difference ocean
models. This approach enables the computational science (performance)
related aspects of a computer model to be kept separate from the
natural (oceanographic) science aspects. In this work we have shown
that the \psykal approach has the potential to improve performance and
to improve code structure/maintainability.

The application of code transformations to the middle/PSy layer is key
to the performance of the \psykal version of a code.

Some of our findings have been sent to the IBM compiler team and will
serve to improve future releases of the compiler.

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}).

@@

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{R.~Ford}
Biography text here.
\end{IEEEbiographynophoto}

@@

\begin{IEEEbiographynophoto}{G.~Riley}
Biography text here.
\end{IEEEbiographynophoto}

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes}
%
% author names and IEEE memberships
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{ A.~R.~Porter, M.~Ashworth, R.~W.~Ford, M.~Modani, G.~D.~Riley% <-this % stops a space
\thanks{A.~R.~Porter, M.~Ashworth, R.~W.~Ford and M.~Modani work in the Science and Technology Facilities Council's Daresbury Laboratory, UK.}% <-this % stops a space
\thanks{G.~D.~Riley works in the School of Computer Science at the University of Manchester, Manchester, UK.}
}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique called \psykal to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion etc) are then
applied to the \psykal version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem sizes
apart from GNU on Haswell, one problem size for Intel on Haswell and
IBM on Power8.
%
As the GNU compiler performs well on Ivybridge the reduction in
performance for the \psykal version was not analysed further. As the
Intel performance on Haswell was better for the \psykal version than
the original version, it was not analysed further.
%
In the case of the IBM compiler, the \psykal performance is
significantly better than the original hand optimised code on all
problem sizes. Analysing the performance of the IBM compiler shows
that a different structure for the original optimised code produces
much improved performance. Replicating this different structure using
appropriate transformations in the \psykal version provides the
same performance improvements.
%
Therefore, the \psykal approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the PSyKAl version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

\end{abstract}

@@

\subsection{The \psykal Approach}

%Proliferation of lightweight cores is resulting in increased
%complexity in the programming models required to achieve performance.
%Combine with the complexity of e.g. a model simulating circulation in
%the global ocean. Significant challenge in software engineering.
%Propose an approach for meeting this challenge by separating the
%computational science demands from the {\it e.g.} oceanographic
%demands.

%\begin{figure}
%\centering
%\caption{The \psykal Approach.}
%\end{figure}

The \psykal approach separates code into three layers; the Algorithm
layer, the PSy layer and the Kernel layer. While this approach is
general, we are currently applying it to Atmosphere and Ocean models
written in Fortran where domain decomposition is typically performed
in the latitude-longitude direction, leaving columns of elements on
each domain-decomposed partition.

The top layer, in terms of calling hierarchy, is the Algorithm
layer. This layer specifies the algorithm that the scientist would like
to perform (in terms of calls to kernel and infrastructure routines)
and logically operates on full fields. We say logically here as the
fields may be domain decomposed, however the algorithm layer is not
aware of this. It is the scientist's responsibility to write this
algorithm layer.

The bottom layer, in terms of calling hierarchy, is the Kernel
layer. The Kernel layer implements the science that the Algorithm
layer calls, as a set of subroutines. These kernels operate on local
fields (a set of elements, a single column of elements, or a set of
columns, depending on the kernel). Again the scientist is responsible
for writing this layer and there is no parallelism specified here,
but, depending on the complexity of the Kernels, there may be input
from an HPC expert and/or some coding rules to help ensure the kernels
compile into efficient code.

The PSy layer sits in-between the Algorithm and Kernel layers and its
functional role is to link the algorithm calls to the associated
kernel subroutines. As the Algorithm layer works on logically global
fields and Kernel layer works on local fields, the PSy layer is
responsible for iterating over columns. It is also responsible for
including any required distributed-memory operations, such as halo
swaps and reductions.

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be optimised and parallelised. The PSy layer
can therefore be tailored for a particular hardware (such as
multi-core, many-core, GPGPUs, or some combination thereof) and
software (such as compiler, operating system, MPI implementation)
configuration with no change to the algorithm or kernel layer
code. This approach therefore offers the potential for portable
performance.

Clearly the separation of code into distinct layers may have an effect
on performance. This overhead, and how to get back to the performance
of a hand optimised code, and potentially improve on it, will be
discussed in the remainder of this paper.

\subsection{The `Shallow' Program}

For this work we use a benchmark called Shallow which solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  In describing
our work, we term the version of the Shallow program obtained at the
beginning of this project the `original' version.

@@

%{\bf Auto-generated BC code by reasoning about meta-data. But then,
%  that brings us on to meta-data...}

@@

The primary aim of this work is to determine whether any performance
loss due to the \psykal separation of concerns can be negated by
optimising the PSy layer. To evaluate this we perform tests on a range
of hardware and a range of compilers. The hardware and compilers are
listed in Table~\ref{TABLE_compilers}. Where a compiler is available
on a given piece of hardware, the version number used in this work is
specified.
%
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
'chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
3.3~GHz.

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
IBM Power 8      &       &             &         & 15.1.2     \\
\hline
\end{tabular}
\end{table}

@@

{\bf RF: Explain that we gradually apply transformations which restructure the psykal code to end up with the same optimised structure as the original code. Explain the reason behind the ordering and what each does - I think you said it was the order that you thought would make the biggest difference. If you can base it on performance analysis then all the better but I don't think that is required.}

@@

The main conclusion of all this analysis is that while we can employ a
separation of concerns and recover performance, doing so is not
straightforward. Determining how best to optimise even a simple code
such as Shallow is highly compiler- and CPU-dependent. Decisions that
are good on one platform may be bad for another and if these are
written into the source code then they will accumulate over time and
will almost certainly result in a code that does not perform optimally
on any system.

{\bf RF: IBM compiler an example of the above point}

@@

We have investigated the application of the \psykal separation of
concerns approach to the domain of finite-difference ocean
models. This approach enables the computational science (performance)
related aspects of a computer model to be kept separate from the
natural (oceanographic) science aspects.

As expected, applying this separation of concerns does reduce
performance significantly when comparing with an existing optimised
code. However, the application of code transformations to the
performance/PSy layer and the use of appropriate compiler flags can
recover any performance losses to within a few percent and in some
cases, despite limiting ourselves to transformations which replicate
the structure of the optimised code, result in slightly improved
performance.

The IBM results demonstrate two key points.
%
Firstly, even for a relatively simple
benchmark code, the code structure required to obtain good performance
between different architecture and/or compiler combinations may
differ. For hand optimised codes this implies the need to support
multiple code bases in order to achieve portable performance.
%
Secondly, the \psykal approach directly addresses the problem of
requiring different optimisations for different architectures and/or
compilers by limiting these differences to a separate performance
layer, thereby allowing the natural science code base to remain
unchanged.

In future work we will extend our performance portability analysis of
the \psykal approach to shared memory parallelisation optimisations on
different architectures using different parallalisation techniques (in
particular OpenMP on multi/many-core and OpenACC on GPU's).
%
We will then analyse the performance of a domain specific compiler
that is currently under development. This compiler will generate
optimised PSy-layer code using a user-provided recipe of
transformations, thereby removing the need for optimisation experts to
manually write the PSy layer.

% \bf{Perhaps we don't want to give too much away here so I've removed the optimisation space search plans even though it is mentioned in the text}
% Finally we plan to use this compiler to allow us to search the optimisation sp%ace of transformations to automatically optimise a code for a particular compil%er/architecture combination or to determine whether we can find improved perfor%mance over existing hand optimised solutions and to automatically

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}). This work was funded by NERC project {\bf xxx}

@@

\begin{IEEEbiographynophoto}{M.~Ashworth}
Biography text here.
\end{IEEEbiographynophoto}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{R.~W~Ford}
Biography text here.
\end{IEEEbiographynophoto}

@@

\begin{IEEEbiographynophoto}{M.~Modani}
Biography text here.
\end{IEEEbiographynophoto}

\begin{IEEEbiographynophoto}{G.~D.~Riley}
Biography text here.
\end{IEEEbiographynophoto}
